Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-12-28 manuscript/manuscript.Rmd
added a few more recent references.

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** @ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape. Achieve handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

**Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing. **Nada Khreisheh**: Investigation, Writing -- review & editing. **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing. **Justin Pargeter**: Conceptualization, Investigation, Methodology, Supervision, Writing -- original draft, Writing -- review & editing.

>>>>

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape. Achieve handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

**Cheng Liu**: Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing. **Nada Khreisheh**: Investigation, Writing -- review & editing. **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing. **Justin Pargeter**: Conceptualization, Investigation, Methodology, Supervision, Writing -- original draft, Writing -- review & editing.
